Viscous Transport in Eroding Porous Media

### 2019-09-30 docs/transport/erosion_transport.tex
small changes to intro and abstract

\begin{abstract} 
  We extend our previous work, [{\em Journal of Computational Physics},
  {\bf 375}:1--21, 2018], to simulate fluid-mechanical erosion of
  two-dimensional porous media with nearly-touching grains.  The method
  continues to use a boundary integral equation (BIE) method to solve
  the fluid equations and to compute the erosion rate, and a stable
  second-order time stepping method to erode the grains.  In our
  original work, we required that the eroding grains were sufficiently
  separated to guarantee that the trapezoid rule provides a sufficient
  amount of accuracy. We now remove this restriction by introducing a
  Barycentric quadrature rule whose error is bounded independent of the
  geometry.  The Barycentric quadrature rule is computationally
  expensive, so it is only used when necessary, and the spectrally
  accurate trapezoid rule is used otherwise. We also simulate and
  analyze transport through eroded porous geometries.  By using the
  Barycentric quadrature rule, the analysis examines flow both in the
  fast and slow regions with high-order accuracy. Our results
  demonstrate the effects of erosion on tortuosity, dispersion, and pore
  size distributions.
\end{abstract}

@@

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterizing
transport through the resulting geometries. Our modeling efforts build
on previous work~\cite{ris-moo-chi-she-zha2012, moore2013self, moo2017},
in particular recent numerical methods developed to simulate erosion in
the regime of Stokes flow~\cite{qua-moo2018}. We, however, make a few
key improvements to the numerical methods from~\cite{qua-moo2018} to
enable simulations of more realistic, dense suspensions of eroding
grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport in the resulting geometries, examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Once streamlines in an eroded geometry are computed, we characterize the
transport by computing and analyzing three different metrics: the
tortuosity, the anomalous dispersion, and the pore size distribution.
The local tortuosity of a streamline connecting the inlet to the outlet
is the streamline's length normalized by the distance between the inlet
and the outlet.  In porous media, depending on several factors including
the porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or even
2~\cite{dud-koz-mat2011}.  The tortuosity of a geometry is defined by
averaging the local tortuosity over all streamlines that are initialized
at the inlet, and the geometry's tortuosity can be used to characterize
average particle motions~\cite{hak-com-den2019}.  To characterize
spreading, the fluid dispersion is computed by calculating the variance
of the streamline lengths. In porous media, this spreading can often be
characterized with a super-diffusive anomalous dispersion
rate~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. An accurate solution near the grains is
crucial since the anomalous dispersion is caused by streamlines spending
time in both the high and low velocity regimes~\cite{ber-sch2001}.
Finally, the third metric, pore size distribution, is useful to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the work of Barnett et
al.~\cite{bar-wu-vee2015} so that high-order accurate derivatives of
solutions of the Stokes equations can be computed.  Using this new
quadrature rule, we accurately compute erosion rates and stably simulate
erosion.  Moreover, we use the quadrature method to visualize the
vorticity in the fluid bulk.  The vorticity is convenient to visualize
since, in the zero Reynolds number limit, it is equivalent to the shear
stress on the eroding bodies, and therefore its magnitude is the erosion
rate.

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are used to characterize the
transport by computing the local tortuosity of trajectories, and the
tortuosity and dispersion rate of eroded geometries. By using the same
quadrature rule and a fourth-order Runge-Kutta method, we are able to
accurately compute trajectories that spend large amounts of time in the
slow velocity regions near the eroding bodies, and reliably computing
these trajectories is essential to characterize the
flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare our tortuosity and
dispersion rates with models available in the literature.  Finally, we
analyze the effect of erosion on the pore size distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work does not have a limitation
on the proximity of the grains. Compared to our previous work, we are
able to simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the model that we
describe in detail in our previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method to compute the shear stress and vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
summarize the model, and a more detailed description is in our previous
work~\cite{qua-moo2018}.  We consider flows inside a confined geometry
$\Omega$ that contains $M$ eroding bodies with boundaries $\gamma_\ell$,
$\ell = 1,\ldots,M$.  The boundary of the bounding geometry is $\Gamma$,
so the boundary of the fluid domain is $\bd \Omega = \Gamma \cup
\gamma_1 \cup \cdots \cup \gamma_M$.  In all examples, we let $\Gamma$
be a slightly rounded-off version of $[-3,3] \times [-1,1]$, and we only
place eroding bodies in the region $[-1,1] \times [-1,1]$.  On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = 0, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.  The strength of $\UU$ is adjusted at
each time step so that there is a constant pressure drop across the
channel.  We simulate erosion by alternating between solving the fluid
equations and advancing the grains according to the erosion rate for a
small time step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require a method to solve the incompressible Stokes equations in
complex two-dimensional porous geometries, we reformulated the
incompressible Stokes equations as a boundary integral equation. This
has the advantage that only the one-dimensional boundary of the domain
must be discretized, and if appropriate quadrature formula and fast
summation methods are used, the result is a high-fidelity numerical
simulation that is computed with near-optimal complexity.

>>>>

\begin{abstract} 
  We extend our previous work, [{\em Journal of Computational Physics},
  {\bf 375}:1--21, 2018], to simulate fluid-mechanical erosion of
  two-dimensional porous media with nearly-touching grains.  The method
  continues to use a boundary integral equation (BIE) method to solve
  the fluid equations, and a stable second-order time stepping method to
  erode the grains.  In our previous work, we required that the eroding
  grains were sufficiently separated to guarantee that the trapezoid
  rule provides a sufficient amount of accuracy. We now remove this
  restriction by introducing a Barycentric quadrature rule whose error
  is bounded independent of the geometry.  The Barycentric quadrature
  rule is computationally expensive, so it is only used when necessary,
  and the spectrally accurate fast-multipole-accelerated trapezoid rule
  is used otherwise. We also simulate and analyze transport through
  eroded porous media.  By using the Barycentric quadrature rule, the
  analysis examines flow both in the fast and slow regions with
  high-order accuracy. Our results reveal several interesting  effects
  of erosion on the tortuosity, dispersion, and pore size distributions.
\end{abstract}

@@

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterizing
transport through the resulting geometries. Our modeling efforts build
on previous work~\cite{ris-moo-chi-she-zha2012, moore2013self, moo2017},
in particular recent numerical methods developed to simulate erosion in
the regime of Stokes flow~\cite{qua-moo2018}. We, however, make a few
key improvements to the numerical methods from~\cite{qua-moo2018} to
enable simulations of more realistic, dense suspensions of eroding
grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport in the resulting geometries, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Once streamlines in an eroded geometry are computed, we characterize the
transport by computing and analyzing three different metrics: the
tortuosity, the anomalous dispersion, and the pore size distribution.
The local tortuosity of a streamline connecting the inlet to the outlet
is the streamline's length normalized by the distance between the inlet
and the outlet.  In porous media, depending on several factors including
the porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or even
2~\cite{dud-koz-mat2011}.  The tortuosity of a geometry is defined by
averaging the local tortuosity over all streamlines that are initialized
at the inlet, and the geometry's tortuosity characterizes average
particle motions~\cite{hak-com-den2019}.  To characterize spreading, the
fluid dispersion is computed by calculating the variance of the
streamline lengths. In porous media, this spreading is often
super-dispersive~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. Accurately resolving the streamlines near
the grains is crucial since the anomalous dispersion is caused by
streamlines spending time in both the high and low velocity
regimes~\cite{ber-sch2001}.  Finally, since they are required to
quantify velocity distributions~\cite{ali-par-wei-bre2017,
dea-qua-bir-jua2018}, channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}, we form the
pore size distribution throughout the erosion process.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the Barycentric quadrature rule of
Barnett et al.~\cite{bar-wu-vee2015} so that high-order accurate
derivatives of solutions of the Stokes equations can be computed.  Using
this new quadrature rule, we accurately compute erosion rates and stably
simulate erosion.  Moreover, we use the quadrature method to visualize
the vorticity in the fluid bulk.

Our second contribution is computing and analyzing trajectories in
eroded geometries.  These trajectories are used to characterize the
transport by computing the local tortuosity of trajectories, and the
tortuosity and dispersion rate of eroded geometries. By using the
Barycentric quadrature rule and a fourth-order Runge-Kutta method, we
accurately compute trajectories that spend large amounts of time in the
slow velocity regions near the eroding bodies, a necessity to
characterize the flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare
our tortuosity and dispersion rates with models available in the
literature.  Finally, we analyze the effect of erosion on the pore size
distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this work is not limited by the
proximity of the grains. Compared to our previous work, we are able to
simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the erosion model that we
describe in detail in our previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method to compute the shear stress and vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
summarize the model, and a more detailed description is in our previous
work~\cite{qua-moo2018}.  We consider flows inside a confined geometry
$\Omega$ that contains $M$ eroding bodies with boundaries $\gamma_\ell$,
$\ell = 1,\ldots,M$.  The boundary of the bounding geometry is $\Gamma$,
so the boundary of the fluid domain is $\bd \Omega = \Gamma \cup
\gamma_1 \cup \cdots \cup \gamma_M$.  In all examples, we let $\Gamma$
be a slightly smoothed version of $[-3,3] \times [-1,1]$, and we only
place eroding bodies in the region $[-1,1] \times [-1,1]$.  On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.  The strength of $\UU$ is adjusted at
each time step so that there is a constant pressure drop across the
channel.  We simulate erosion by alternating between solving the fluid
equations and advancing the grains according to the erosion rate for a
small time step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since we require an accurate solution of the incompressible Stokes
equations in complex two-dimensional porous media, we reformulated the
incompressible Stokes equations as a BIE.  This has the advantage that
only the one-dimensional boundary of the domain must be discretized, and
if appropriate quadrature formula and fast summation methods are used,
the result is a high-fidelity numerical simulation that is computed with
near-optimal complexity.
